Vragenlijst thesis evaluatie

### 2015-06-07 doc/template/chap-10.tex
aanvullingen tekst

The game theoretical approach of the FlipIt does not take such delay into account. 
This an lead us to the following research questions:
\begin{itemize}
\item Is it possible to incorporate the notion of delay in the game-theoretical analysis of the Flip-It game ?
\item Does this allow us to determine an optimal defense strategy against an attacker ? for example: Gaat er een specieke grootte zijn van een delay waarbij de attacker al weet dat hem niet meer moet gaan spelen ? ( is niet gelijk aan de grootte van de periode van de attacker)
\end{itemize}

Resarch questions when working with a network and a delay: .. graph model en uitleggen hoe we de graph kunnen maken zodat de delay altijd zo groot mogelijk gaat zijn. 
\begin{itemize}
\item How can we calculate the expected duration for a node's infection/the entire network infection ?
\item Can we calculate this node per node ?
\end{itemize}

>>>>

The game theoretical approach of the FlipIt does not take such delay into account. 
This an lead us to the following research questions:
\begin{itemize}
\item How can we incorporate the notion of delay in the game-theoretical analysis of the Flip-It game for a periodic strategy ?
\item Is there an optimal defense strategy against an attacker ? 
\end{itemize}

### 2015-06-07 doc/template/chap-5.tex
abstract
aanvullingen tekst

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%			Introduction Chapter 1				%%%%%%
%%%%%												%%%%%%
%%%%%												%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Gametheory is a mathematical study to analyse interactions between independent and self-interested agents. To get an understanding of the most important concepts of game theory, a short introduction based on the work of 
\cite{leyton2008essentials} and \cite{Coursera} is given in section \ref{Cha1:briefintro}  . For a more detailed and full introduction to game theory, the reader is referred to 
\cite{leyton2008essentials}.  In section \ref{cha1:FlipItGame} an overview of the FlipIt game is given with the definitions and concepts that will be used throughout the paper. 
The last section \ref{ch1:extendedWork} will cover the extensions and additions already made on FlipIt.

@@

@@

The state of the resource is denoted as a time-dependent variable $C=C_{i}(t)$. 
$C_{D}(t)$ is 1 if the game is under control by the defender and 0 if the game is under control by the attacker. Reversely, $C_{A}(t)$ is 1 if the game is under control by the attacker and 0 if under control by the defender. So, $C_{A}(t)= 1 - C_{D}(t)$.
The game starts with the defender being in control: $C_{D}(0)= 1$. \\

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%			Introduction Chapter 1				%%%%%%
%%%%%												%%%%%%
%%%%%												%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What is cyber security?}

Cyber security is a part of security that focuses on securing computer controlled devices such as computers and smartphones, as well as public and private computer networks, including the entire Internet. A private network can be an enterprise network that is protected from the public network such as the internet.  The purpose of security is to give certainty that data will not be removed without authorization (Confidentiality) that the data is always accessible (Availability), and that data can not be read or altered by someone who does not have the authorization (Integrity). These are the 3 key attributes of security, also known as the CIA triad. \\

In order to keep a system secure, it is important to mitigate the possible threats. A threat refers to anything that can breach the security and that can cause a possible harm. The some of the most common threats are spam, malware, spoofing, phishing and  DDoS attacks. 
The biggest threat are caused by malware. 
This is a piece of malicious program that is designed to penetrate unprotected systems or computers, and getting there to sensitive information. Malware can be divided into different subclasses. These subclasses are not mutually exclusive:

\begin{description}
\item Virus: This is a malicious piece of code that replicates and tries to spread to infect other systems or files. The "I love you" virus is an example of a virus that quickly spreads. This virus propagates via mail systems. If someone opens an email with "I love you" with the virus in the annex this virus spreads itself by sending a mail to everyone in your contact list. So the virus can multiply rapidly and eventually a business network might shut down by the heavy traffic. In this example, there is a need for human interaction to spread the virus. If no one opens the mail the virus can not infect other systems.
Unfortunately, there are also viruses that can spread without human interaction. These viruses are referred to as worms. A worm is also a computer program that replicates itself to spread to other computers. Via a computer network, copies of the worm can be forwarded without an intermediary. The worm will use vulnerabilities of the system to infect other computers.
The Stuxnetworm is a very famous worm. Initially this worm spread via infected USB sticks and from then it could spread through the Internet to other computers. The purpose of the Stuxnetworm was to harm the centrifuges in nuclear reactors. Many reactors have been infected. From the standpoint of the defender, it is very important to respond as quickly as possible so that the worm can not spread. 
\item Rootkit:
\item Trojan:
\item Backdoors:
\end{description}

A security report of 2014 reveales that 61\% of the attacks on companies are caused by malware. A company can take different measures to defend himself against malware. 
The threats caused by malware can be devided into 3 big groups: known threats (70\%), unknown threats (29\%) and advanced threats (1\%) [\todo{toevoegen referentie naar kaspersky APT report}].

The known threats easy to defend against. Unkown also. Then the remaining 1\% are the advanced threats. These threats are also known as APT. APT's are ... 
In this paper we focus on targeted attacks. Targeted attacks are malware designed to attack exclusively an organisation. Some examples of the biggest most rare APTs  to get an understanding of what APTs are capable of and how long they can stay unseen. [\todo{site kaspersky apt}]

An APT is a persistent targeted attack that tries to penetrate a network, harm and stays unseen for a long period of time. The motive of an APT is mostly cyber espionage, stealing sensitive data, sabotage or other ideological attacks. APT stands for Advanced persistant threat. Advanced stands for the fact that these attacks are well funded and that the attacker itself has a great expertise. These APT are not easy to protect against. Not all APTs are technical advanced. The attacker can also try to exploit existing vulnerabilities and hope that his target has not yet secured himself against these vulnerabilities. Persistent is the fact that the attacker keeps on trying to break in to his victim. He will not give up. The attack can be over various years and different steps. The threats stands for the fact that an APT is ..

\subsubsection{Equation}
Equation is a complex cyberattack platform where the first known sample is from 2002, but it was only discovered 12 years later in 2014. This APT propagates through usb drives, cd or physical media. It will search for exploits and will self-replicate itself to spread the infection. The purpose of this virus is to steal data and cyberespionage.

\subsubsection{Regin}

\subsubsection{Flame}
Way of propagation through USB drives, LAN spreading. purpose cyber espionage.

\subsubsection{Black energy}
 purpose cyber espionage and DDoS, data wiping. prop usb lan

According to [\todo{APT report} ]  the damage of one successful target attack can exceed over 2.54 million dollar. a company needs a defense mechanism to defend itself against apt. Prevention more important than ..
een van de manieren om dat te doen is via game theorie.

Security belangrijk. verschillende malware. uitleggen malware. verwijzen naar kaspersky security raport. APT grote threats. voorbeelden van APT's. waarom gametheory gebruiken.

\section{Intro game theory}
Gametheory is a mathematical study to analyse interactions between independent and self-interested agents. To get an understanding of the most important concepts of game theory, a short introduction based on the work of 
\cite{leyton2008essentials} and \cite{Coursera} is given in section \ref{Cha1:briefintro}  . For a more detailed and full introduction to game theory, the reader is referred to 
\cite{leyton2008essentials}.  In section \ref{cha1:FlipItGame} an overview of the FlipIt game is given with the definitions and concepts that will be used throughout the paper. 
The last section \ref{ch1:extendedWork} will cover the extensions and additions already made on FlipIt.

@@

 \todo{optimal strategy uitleggen}

@@

We denote the state of the resource as a time-dependent variable $C=C_{i}(t)$. 
$C_{D}(t)$ is 1 if the game is under control by the defender and 0 if the game is under control by the attacker. Reversely, $C_{A}(t)$ is 1 if the game is under control by the attacker and 0 if under control by the defender. So, $C_{A}(t)= 1 - C_{D}(t)$.
The game starts with the defender being in control: $C_{D}(0)= 1$. \\

### 2015-06-07 doc/template/chap-9.tex
aanvullingen tekst

\begin{description}
\item $\delta_{i}$
\item 
\end{description}

@@

@@

\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) =\dfrac {r} {2} - k_{A} \alpha_{A} = \dfrac {\delta_{D}} {2\delta_{A}} - k_{A} \alpha_{A}  
\end{equation}\\

Correspondingly, the benefit for the defender can be expressed as:
\begin{equation}\label{first}
\beta_{D}(\alpha_{D},\alpha_{A}) =1 -  \dfrac {r} {2} - k_{D} \alpha_{D} = 1 - \dfrac {\delta_{D}} {2\delta_{A}} - k_{D} \alpha_{D} 
\end{equation}

@@

However, because of the delay required for virus propagation, the maximal time of control is reduced to $\delta_{D}-d$ , see figure \ref{fig:delaycase1}. There is a probability of \textit{r} that the attacker will move in the interval of the defender. However, the gain will not be half of the interval. Indeed, the attacker has to play soon enough to gain control, meaning that the attacker has to play during the period of $\delta_{D}-d$ during the interval of the defender. The probability that the attacker plays soon enough is $\dfrac{\delta_{D}-d}{\delta_{D}}$ and this will give the attacker an average gain of $\dfrac{\delta_{D}-d}{2}$. If the attacker moves after the period of $\delta_{D}-d$, the gain of the attacker will be zero. The probability that this happens is  $\dfrac{d}{\delta_{D}}$. The average gain rate of the attacker can then be expressed as follows if we look at one interval of the defender:
\begin{equation}\label{first}
\gamma_{A}(\alpha_{D},\alpha_{A}) = \dfrac {1}{\delta_{D}} [ \dfrac{\delta_{D}}{\delta_{A}} \cdot \dfrac{\delta_{D}-d}{\delta_{D}} \cdot \dfrac{\delta_{D}-d}{2} + \dfrac{\delta_{D}}{\delta_{A}} \cdot \dfrac{d}{\delta_{D}} \cdot 0 ]
\end{equation}

To derive the benefit, the cost of moving is subtracted from the average gain. 
\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) = \dfrac { (\delta_{D}-d) ^{2}} {2 \cdot \delta_{D}  \delta_{A}} - k_{A} \alpha_{A}
\end{equation}
\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) = \dfrac { \delta_{D}} {2 \cdot \delta_{A}} - k_{A} \alpha_{A} - ( \dfrac{d^{2}}{2 \cdot \delta_{A} \delta_{D}} - \dfrac{d}{\delta_{A}} )
\end{equation}

 The benefit of the defender is expressed as follows:
 \begin{equation}\label{first}
\beta_{D}(\alpha_{D},\alpha_{A}) = 1 - \dfrac { (\delta_{D}-d) ^{2}} {2 \cdot \delta_{D}  \delta_{A}} - k_{D} \alpha_{D}
\end{equation}
~~\\
We can easily see that when $d$=0, we obtain the formula of the original FlipIt game.\\

@@

\begin{equation}\label{first}
\beta_{D}(\alpha_{D},\alpha_{A}) = \dfrac {1} {2r} - k_{D} \alpha_{D} = \dfrac {\delta_{A}} {2\delta_{D}} - k_{D} \alpha_{D} 
\end{equation}
\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) =1 - \dfrac {1} {2r} - k_{A} \alpha_{A} = 1- \dfrac {\delta_{A}} {2\delta_{D}} - k_{A} \alpha_{A}  
\end{equation}\\

@@

\subsubsection*{\textbf{Case a:} $d + \delta_{A} \leq \delta_{D}$}
~~\\
Consider a timespan $\delta_{A} + d$, representing the attacker's interval followed by the delay period in his next interval. The defender will never move twice during this timespan because $\delta_{A} + d \leq \delta_{D}$. The defender will move during the interval of the attacker with a probability of $\dfrac{\delta_{A}}{\delta_{D}} $. When this happens the defender will end with being in control at the end of the interval. In the next interval the attacker will have to regain control, meaning that during the delay, the defender stays in control, see figure \ref{fig:case2} cases (1) and (2). This means that the defender will keep the control over the resource in the next interval over a period of the delay, namely \textit{d}. Because $d + \delta_{A} \leq \delta_{D}$ the next move of the defender in this second interval will never occur during the delay, meaning that the entire delay can be considered as an extra benefit resulting of a play in the previous interval. 
So, every time the defender plays, he will get an average gain of $\dfrac{\delta_{A}}{2}$ in the interval where he plays and in the next interval will always receive a extra gain of $d$, yielding a total average gain per interval of
$\dfrac{(d+\dfrac{\delta_{A}}{2})}{\delta_{A}}$

>>>>

\begin{description}
\item $i$: Defines the player. Different as in the FlipIt paper where the defender is denoted by the subscript 0 and the attacker by the subscript 1.
\item $\delta_{i}$: The length of the interval between two consecutive moves of player \textit{i}. 
\item $\alpha_{i}$: The average flip rate of player \textit{i}, given by $\alpha_{i}=1/\alpha_{i}$.
\item $k_{i}$: The cost of player \textit{i}'s moves.
\item $d$: The delay caused by the virus propagation.
\item $G_{i}(t)$: The total gain of player \textit{i} denotes the amount of time player \textit{i} is in control over the resource up to time \textit{t}.
\item $\gamma_{i}$: The average gain rate of player \textit{i} defined as $G_{i}(t)/t$
\item $\beta_{i}$:  The average benefit rate up to time \textit{t} defined as  $\beta_{i} = \gamma_{i} -k_{i} \alpha_{i} $.
\item $opt_{i}$: The optimum function.
\end{description}

\todo{tekening toevoegen waarop de delay staat aangeduid}

@@

Intuitively we could assume that $d$ can never be bigger than $\delta_{A}$ because then the benefit for the defender would always be 1. This is not always true. It is only true if $d$ is bigger than $\delta_{D}$, because then the defender will always be in control. For this we only calculate the formulas for the cases where $d$ is smaller than $\delta_{D}$. We can already conclude that it is no use for the attacker to play when the delay is bigger than $\delta_{D}$. \\

@@

\begin{equation}
\beta_{A}(\alpha_{D},\alpha_{A}) =\dfrac {r} {2} - k_{A} \alpha_{A} = \dfrac {\delta_{D}} {2\delta_{A}} - k_{A} \alpha_{A}  
\end{equation}\\

Correspondingly, the benefit for the defender can be expressed as:
\begin{equation}
\beta_{D}(\alpha_{D},\alpha_{A}) =1 -  \dfrac {r} {2} - k_{D} \alpha_{D} = 1 - \dfrac {\delta_{D}} {2\delta_{A}} - k_{D} \alpha_{D} 
\end{equation}

@@

However, because of the delay required for virus propagation, the maximal time of control is reduced to $\delta_{D}-d$ , see figure \ref{fig:delaycase1}. There is a probability of \textit{r} that the attacker will move in the interval of the defender. However, the gain will not be half of the interval. Indeed, the attacker has to play soon enough to gain control, meaning that the attacker has to play during the period of $\delta_{D}-d$ during the interval of the defender. The probability that the attacker plays soon enough is $\dfrac{\delta_{D}-d}{\delta_{D}}$ and this will give the attacker an average gain of $\dfrac{\delta_{D}-d}{2}$. If the attacker moves after the period of $\delta_{D}-d$, the gain of the attacker will be zero. The probability that this happens is  $\dfrac{d}{\delta_{D}}$. The average gain rate of the attacker can then be expressed as follows if we look at one interval of the defender:
\begin{equation}
\gamma_{A}(\alpha_{D},\alpha_{A}) = \dfrac {1}{\delta_{D}} [ \dfrac{\delta_{D}}{\delta_{A}} \cdot \dfrac{\delta_{D}-d}{\delta_{D}} \cdot \dfrac{\delta_{D}-d}{2} + \dfrac{\delta_{D}}{\delta_{A}} \cdot \dfrac{d}{\delta_{D}} \cdot 0 ]
\end{equation}

To derive the benefit, the cost of moving is subtracted from the average gain. 
\begin{equation}
\beta_{A}(\alpha_{D},\alpha_{A}) = \dfrac { (\delta_{D}-d) ^{2}} {2 \cdot \delta_{D}  \delta_{A}} - k_{A} \alpha_{A}
\end{equation}
\begin{equation}
\beta_{A}(\alpha_{D},\alpha_{A}) = \dfrac { \delta_{D}} {2 \cdot \delta_{A}} - k_{A} \alpha_{A} - ( \dfrac{d^{2}}{2 \cdot \delta_{A} \delta_{D}} - \dfrac{d}{\delta_{A}} )
\end{equation}

 The benefit of the defender is expressed as follows:
\begin{equation}
\beta_{D}(\alpha_{D},\alpha_{A}) = 1 - \dfrac { (\delta_{D}-d) ^{2}} {2 \cdot \delta_{D}  \delta_{A}} - k_{D} \alpha_{D}
\end{equation}
~~\\
We can easily see that when $d$=0, we obtain the formula of the original FlipIt game.\\

@@

\begin{equation}
\beta_{D}(\alpha_{D},\alpha_{A}) = \dfrac {1} {2r} - k_{D} \alpha_{D} = \dfrac {\delta_{A}} {2\delta_{D}} - k_{D} \alpha_{D} 
\end{equation}
\begin{equation}
\beta_{A}(\alpha_{D},\alpha_{A}) =1 - \dfrac {1} {2r} - k_{A} \alpha_{A} = 1- \dfrac {\delta_{A}} {2\delta_{D}} - k_{A} \alpha_{A}  
\end{equation}\\

An intuitive solution for the case with a virus would be to subtract the benefit of the attacker received in each interval with the delay similarly as in case 1. This would give us the following formula if we derive it in the same way:
\begin{equation}
\beta_{A}=\dfrac{(\delta_{A} - d)^2}{2\delta_{A}\delta_{D}} - \dfrac{k_{A}}{\delta_{A}}
\end{equation}

But this results in an overestimation. How closer $\delta_{A}/\delta_{D}$ is equal to one, the better the approximation. If $\delta_{A}/\delta_{D} = 1$ the result is correct. This formula does not take into account that if the attacker was in control in the previous interval, the delay should not be subtracted, because the delay will not be in control of the defender. This means that we have to look at what happens in the previous interval. \\

From the defender we know that his moves are instantaneous. It is easier to calculate the benefit of the defender in this case. Because the defender moves slower than the attacker we know that if the defender moves during the interval of the attacker, he only moves once within this interval.
The defender will move during the interval of the attacker with a probability of $\dfrac{\delta_{A}}{\delta_{D}} $. When this happens the defender will end with being in control at the end of the interval. In the next interval the attacker will have to regain control, meaning that during the delay, the defender stays in control, see figure \ref{fig:case2} cases (1) and (2). The defender will keep the control over the resource in the next interval over a period of the delay, namely \textit{d}. \\

Consider a timespan $\delta_{A} + d$, representing the attacker's interval followed by the delay period in his next interval. The defender will never move twice during this timespan because $\delta_{A} + d \leq \delta_{D}$.
Because $d + \delta_{A} \leq \delta_{D}$ the next move of the defender in this second interval will never occur during the delay, meaning that the entire delay can be considered as an extra benefit resulting of a play in the previous interval. 
So, every time the defender plays, he will get an average gain of $\dfrac{\delta_{A}}{2}$ in the interval where he plays and in the next interval will always receive a extra gain of $d$, yielding a total average gain per interval of
$\dfrac{(d+\dfrac{\delta_{A}}{2})}{\delta_{A}}$

@@

\subsubsection*{\textbf{Case a:} $d + \delta_{A} \leq \delta_{D}$}
